---
title: "Question sets-Uniform distribution and empirical distribution"
format:
  revealjs:
    output-file: 1-Question sets_presentation.html
    theme: [default, student_view.scss]
    slide-number: true
    reference-location: margin
    chalkboard: 
      theme: whiteboard
format-links: true
engine: knitr 
bibliography: ref.bib
---

## Question 1 {.smaller}

Given $X \sim \text{Uniform}(a, b)$, 

1. Please show that $\text{E}(X) = \frac{a+b}{2}$

:::{.hide}
$$
\begin{align*}
\mathbb{E}(X) &= \int_{-\infty}^{\infty} x\, f_X(x)\, dx 
= \int_{a}^{b} x \cdot \frac{1}{b-a}\, dx 
= \frac{1}{b-a} \int_{a}^{b} x\, dx \\
&= \frac{1}{b-a} \left[ \frac{x^2}{2} \right]_{a}^{b} 
= \frac{1}{b-a} \left( \frac{b^2}{2} - \frac{a^2}{2} \right) 
= \frac{b^2 - a^2}{2(b-a)} 
= \frac{(b-a)(b+a)}{2(b-a)} \\[6pt]
&= \frac{a+b}{2}
\end{align*}
$$
:::

## {.smaller}

2. Please show that $\text{VAR}(X) = \frac{(b-a)^2}{12}$

:::{.hide}
$$
\begin{align*}
\mathrm{Var}(X) &= \int_a^b \left( x - \frac{1}{2} (a+b) \right)^2 \cdot \frac{1}{b-a} \, \mathrm{d}x 
= \frac{1}{b-a} \cdot \int_a^b \left( x - \frac{a+b}{2} \right)^2 \, \mathrm{d}x \\
& \text{let } t= x - \frac{1}{2} (a+b) \text{ and } h= \frac{1}{2} (b-a) \\
&= \frac{1}{b-a} \int^h_{-h}t^2 = \frac{1}{3(b-a)} \cdot [ t^3 ]_{-h}^h \\
&= \frac{1}{3(b-a)}(\frac{1}{2} (b-a))^3-(-\frac{1}{2} (b-a))^3 \\
&= \frac{(b-a)^2}{12}
\end{align*}
$$

:::

## {.smaller}

3. Please show that the CDF is:
$$  
\begin{aligned}
F(x) \begin{cases}
0
 \quad & \text{if }x < a \\
\frac{x-a}{b-a}
 \quad & \text{if } a \leq x < b \\
1
 \quad & \text{if } x \geq b 
\end{cases} 
\end{aligned}
$$

:::{.hide}

- Case 1: $x < a$
For $t < a$, $f_X(t) = 0$, so the integral from $-\infty$ to $x$ is zero:

- Case 2: $a \leq x < b$
  
$$
\begin{align*}
F(x) &= \int_{-\infty}^{x} f_X(t)\, dt 
     = \int_{-\infty}^{a} 0\, dt + \int_{a}^{x} \frac{1}{b-a}\, dt \\
     &= 0 + \frac{1}{b-a} \int_{a}^{x} 1\, dt 
     = \frac{1}{b-a} \Big[ t \Big]_{a}^{x} \\
     &= \frac{x - a}{b - a}.
\end{align*}
$$

- Case 3: $x \geq b$
   Similar to Case 1

:::

## {.smaller}

4. Please show that $\mathbb{E}[(X - \mathbb{E}(X))^3] = 0$


:::{.hide}
$$
\begin{align*}
\mathbb{E}(X-\mathbb{E}(X))^3  &= \int_a^b \left( x - \frac{1}{2} (a+b) \right)^3 \cdot \frac{1}{b-a} \, \mathrm{d}x 
= \frac{1}{b-a} \cdot \int_a^b \left( x - \frac{a+b}{2} \right)^3 \, \mathrm{d}x \\
& \text{let } t= x - \frac{1}{2} (a+b) \text{ and } h= \frac{1}{2} (b-a) \\
&= \frac{1}{b-a} \int^h_{-h}t^3  \\ 
& \text{because } t^3 \text{ is an odd function, so} \\
&= \frac{1}{b-a} \int^h_{-h}t^3 = 0
\end{align*}
$$
:::


## Question 2 {.smaller}

For a given empirical distribution

1. Please show that the empirical distribution fulfill non-negativity

:::{.hide}

$$  
\begin{aligned}
\because N>0  \quad  \&  \quad 
I(x)\begin{cases}
1
 \quad & \text{if }X\text{ is TRUE} \\
0
 \quad & \text{if }X\text{ is FALSE}
\end{cases} \geq 0 \\
\rightarrow \hat{f}_N(x) = \frac{1}{N} \sum_{i=1}^N I(x_i = x)  \geq 0
\end{aligned}
$$

:::

## {.smaller}

2. Please show that the empirical distribution fulfills the unit measure condition

:::{.hide}

Let the support have $t$ unique values $\{X_1, \dots, X_t\}$:
$$  
P(X) = \sum_{t} \frac{1}{N} \sum_{i=1}^N I(x_i = X_t) = \frac{N}{N}= 1
$$

:::

## Question 3 {.smaller}

Given $X \sim \text{Uniform}(a, b)$, and for any numbers $u, v, w$ where $a < u < v < w < b$ and $v - u = w - v = c$, please show that:

$\text{P} (u\leq X\leq v) = \text{P} (v\leq X\leq w)$

:::{.hide}
$$
\small
\begin{align*}
\text{P} (u\leq X\leq v) &= P(x\leq v)-P(x\leq u) \\ &= \frac{v - a}{b - a}-\frac{u - a}{b - a}=\frac{c}{b-a}\\
\text{P} (v\leq X\leq w) &= P(x\leq w)-P(x\leq v) \\ &= \frac{w - a}{b - a}-\frac{v - a}{b - a}=\frac{c}{b-a}
\end{align*}
$$
:::

## Question 4 {.smaller}

Hoo Hey How (魚蝦蟹) is a traditional Southern Chinese dice game rooted in Hokkien culture and popularly played during festivals like Chinese New Year. Using three six-sided dice marked with symbols—typically fish, prawn, crab, gourd, rooster, and stag. Players place bets on a board featuring these icons. After the dice are rolled, payouts are awarded based on how many times a chosen symbol appears: 1:1 for one match, 2:1 for two, and 3:1 for three. This game is also popular in Vietnam called Bầu Cua Cá Cọp and Cambodia called Klah Klok.

![Illustration of Hoo Hey How By Outlookxp - Own work, CC BY-SA 3.0, [Link](https://commons.wikimedia.org/w/index.php?curid=30465860) ](img\hooHeyHow.JPG)

Suppose that you bet 1 dollar on Crab. Let $X$ denote the money you win (negative value represents a loss) from one trial of this game

1. What is the support of $X$

:::{.hide}
$-1,1,2,3$
:::

## {.smaller}

2. What is the probability mass function (pmf) of X

:::{.hide}
$$  
\begin{aligned}
p(x) \begin{cases}
\frac{125}{216}
 \quad & \text{, when }x = -1 ;\\
\frac{25}{72}
 \quad & \text{, when }x = 1 ;\\
 \frac{5}{72}
 \quad & \text{, when }x = 2 ;\\
 \frac{1}{216}
 \quad & \text{, when }x = 3 ;\\
  0
 \quad & \text{Otherwise.} \\
\end{cases} 
\end{aligned}
$$
:::


---

3. What is the cumulative distribution function (CDF) of X?

:::{.hide}
$$  
\begin{aligned}
F(x) \begin{cases}
0
 \quad & \text{, when }x < -1 ;\\
\frac{125}{216}
 \quad & \text{, when } -1 \leq x < 1 ;\\
 \frac{25}{27}
 \quad & \text{, when } 1 \leq x < 2 ;\\
 \frac{215}{216}
 \quad & \text{, when } 2 \leq x < 3 ;\\
  1
 \quad & \text{, when }  x \geq 3. \\
\end{cases} 
\end{aligned}
$$
::: 

## Question 5 {.smaller}

“Pig” is a simple dice game in which two players take turns to roll a six-sided die, according to the following rule

- If a player rolls a 1, the player scores nothing and it becomes the opponent’s turn.
- If a player rolls any other number, it is added to his turn total and the player’s turn continues.
- If the player instead chooses to hold, the turn total is accumulated to his or her score and
it becomes the opponent’s turn.
- The first player who scores 100 or more points wins

A simple tactic at the early stage of the game is called “hold at k strategy”, with which one should continue to roll whenever the turn total is less than k. What would be the wise choice on the value of k?

## {.smaller}

1. Given the current turn total is $s$ what is the expected gain of rolling 

:::{.hide}
$$  
E(X)=-s\frac{1}{6}+2\times\frac{1}{6}+3\times\frac{1}{6}+4\times\frac{1}{6}+5\times\frac{1}{6}+6\times\frac{1}{6}=+2\times\frac{20-s}{6}
$$
::: 

:::{.callout-note}

1. Holds at k strategy is only good in early game
2. The total optimal strategy actually varies based on your opponent's situation.
3. Detailed discussion can be found in @neller2005pigtail

:::

## Question 6: Birthday Problem {.smaller}

What is the probability that at least two people share the same birthday in the classroom with size of 20 ?

When we examined this problem previously, we assumed:

- 365 days in a year
- Every birthday is equally likely to happen

## {.smaller}
1. Assuming each individual's birthday is independent and equally likely to occur on any day, what is the potential distribution of everyone's birthday?


:::{.hide}
Uniform distribution (Either continuous and discrete is fine in this case)
:::

## {.smaller}
2. What is the probability of all student's birthday are different? 

:::{.hide}
$$
 =0.59
$$
:::

## {.smaller}
3. What is the probability that at least two people share the same birthday in the classroom with size of 20 ?

:::{.hide}
$$
1-\frac{\binom{365}{20}\times 20!}{365^{20}} =\frac{P^{365}_{20}}{365^{20}}=0.41
$$
:::

## {.smaller}
However, these assumptions are unrealistic. Let's refine our model by 
using real world data.

### Data

The UK publishes the average frequency of births on each day of the year
from 1995 to 2024. We can download it with

```
curl -o uk-daily-births.csv \
https://www.ons.gov.uk/visualisations/nesscontent/dvc307/line_chart/data.csv
```
## {.smaller}
So, let's use this data to construct the empirical 
probability mass function (pmf).
Then, we can re-estimate the probability that two people share the same
birthday in a group of people.

### Tasks

1. Load the data into R and visualize it.
   
```{r}
library(tidyverse)
library(lubridate) # Library for handeling datetime 

# 1. Load the data
data <- read_csv("Datasets/data.csv", col_types = cols(
  date = col_character(),
  average = col_double()
))

# Parse the date properly (day-month format, ggplot2 cannot accept date without year information)

data <- data %>%
  mutate(
    date_parsed = dmy(paste(date, "2024")),  
    month = month(date_parsed),
    day   = day(date_parsed),
  )

# Visualise daily average births0
p1 <- ggplot(data, aes(x = date_parsed, y = average)) +
  geom_line(color = "steelblue") +
  geom_point( color = "steelblue") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = "Average Daily Births in England & Wales (1995–2024)",
       x = "Date", y = "Average number of births") +
  theme_minimal()

p1
```

## {.smaller}
2. Construct the empirical pmf that a person is born on each day within a year.

```{r}

# Account for leap years and compute total births per calendar day ----
# Over 30 years (1995–2024): 7 leap years (1996,2000,2004,2008,2012,2016,2020,2024)

data <- data %>% mutate(pmf_weight=average/sum(average) # Total births in the whole period
)
```

```{r}
# Verify it sums to 1
cat("PMF sums to:", sum(data$pmf_weight), "\n")

# Visualise the empirical distribution

ggplot(data, aes(x = date_parsed, y = pmf_weight)) +
  geom_col(fill = "darkorange", alpha = 0.8, width = 1) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = "Empirical Birth Probability by Calendar Day (366-day year)",
       x = "Month", y = "Probability") +
  theme_minimal()
```

## {.smaller}
3. Implement a simulation that generates 20 people their  birthday based on the empirical pmf.


```{r}
inverse_cdf <- function(u,pmf) {
   cdf_366 <- cumsum(pmf)                # Get cdf
   day_indices<-1:366                    # day-of-year 1 … 366
   return(day_indices[which(cdf_366 >= u)[1]])   # first day where CDF exceeds u
}
```

vectorization
```{r}
inverse_cdf_vec <- Vectorize(inverse_cdf,vectorize.args = 'u') # vectorization on u to facilitate speed computation
data$date_parsed[inverse_cdf_vec(runif(20),data$pmf_weight)]
```


## {.smaller}

4. Compare $P\left( X \ge 2 \right)$ determined using the simple uniform model 
   we used previously vs. the probability that is estimated using the 
   empirical pmf.

```{r}
simulation_amount=2000 # Number of simulation

class_size=20 # Number of students in each class (group size)

# Step 1: Generate random birthdays using the INVERSE-CDF method
#         We generate 'simulation_amount × class_size' uniform random numbers U(0,1)
#         Then transform each U into a day-of-year using the empirical distribution

birthdays_matrix <- matrix(
    inverse_cdf_vec(runif(simulation_amount * class_size),data$pmf_weight),   # generate n_sim × n_people uniforms → days
    nrow = simulation_amount,
    ncol = class_size
  )


# Step 2: Check for each simulated classroom whether at least two students
#         share the same birthday
#         We use: if number of unique birthdays < total students → at least two people have the same birthday
check_same_birthsday <- apply(birthdays_matrix,MARGIN=1,function(x) length(unique(x))<length(x))

# Step 3: Estimate the probability
#         Proportion of simulated classrooms that had at least one shared birthday
probability= sum(check_same_birthsday)/simulation_amount
probability
```