## Comparison

| Operator | Description              | Example  | Returns |
|----------|--------------------------|----------|---------|
| `==`     | Equal to                 | `5 == 5` | TRUE    |
| `!=`     | Not equal to             | `5 != 7` | TRUE    |
| `>`      | Greater than             | `10 > 8` | TRUE    |
| `<`      | Less than                | `2 < 4`  | TRUE    |
| `>=`     | Greater than or equal to | `5 >= 5` | TRUE    |
| `<=`     | Less than or equal to    | `3 <= 5` | TRUE    |

## Logical operators

 **Logic operators** are used for operating on logic data type (i.e. TRUE or FALSE). The three logical operators are:

- AND `&`
- OR `|`
- NOT `!`

```{r}
#| echo: true
#| eval: true
#| output-location: column
# To test if Ray will have to lecture on Zoom
zoomClass <- function(blackRain, Typhoon8){
  return (blackRain | Typhoon8)
}

zoomClass(TRUE, TRUE) # It is Black Rain and Typhoon 8
```
```{r}
#| echo: true
#| eval: true
#| output-location: column
zoomClass(FALSE, TRUE) # It is only Typhoon 8
```
```{r}
#| echo: true
#| eval: true
#| output-location: column
zoomClass(TRUE, FALSE) # It is only Black Rain
```
```{r}
#| echo: true
#| eval: true
#| output-location: column
zoomClass(FALSE, FALSE) # It is neither
```


## If/else

Link logical outcome to an action given the condition. This is where the `if` and `else` statement comes in. The format is like

```R
if (CONDITION) {
  # If CONDITION==TRUE: codes inside this bracket will be executed
} else {
 # If CONDITION==FALSE: codes inside this bracket will be executed (You can also omit this part if needed)
}
```
**Example**
```{r}
#| echo: true
#| eval: true
#| output-location: column
# example of if/else based on passing exam
zoomClass <- function(blackRain, Typhoon8){
  if  (blackRain | Typhoon8) {
    print("Zoom link:XXXX")
  }else{
    print("We will have offline class")
  }

}
zoomClass(TRUE, TRUE)
```

```{r}
#| echo: true
#| eval: true
#| output-location: column
zoomClass(FALSE, FALSE)
```

## Chained conditions

It is also possible to chain if/else statements if there are multiple conditions:

```{r}
#| echo: true
#| eval: true
#| output-location: column
# example of multiple if else conditions
ageCategory <- function(age){
  if (age <= 12){
    print("Child")
  } else if (age > 12 & age <= 19){
    print("Teenage")
  } else if (age >= 20 & age <= 60) {
    print("Adult")
  } else {                                # > 60
    print("Elderly")
  }                              
}
```

```{r}
#| echo: true
#| eval: true
#| output-location: column
ageCategory(5)
```

```{r}
#| echo: true
#| eval: true
#| output-location: column
ageCategory(45)
```

```{r}
#| echo: true
#| eval: true
#| output-location: column
ageCategory(70)
```

## Loop

- Using loop to reduce redundant code

Here are some simple examples of a for loop in R:

```{r}
#| echo: true
#| eval: true
#| output-location: column
# examples of loop
# Loop printing a series of numbers from 1 to 10
for (i in 1:10){
  print(i)
}
```
You can also iterate over vectors
```{r}
#| echo: true
#| eval: true
#| output-location: column
# Loop printing a vector of names
names <- c("Jason","Ray","Yuanhua")
for (i in names){
  print(i)
}
```