## Data structures

Data structures are one of the most important features in programming. 

- Organize data
- Store data

**Benefits**
- Easier to read
- Faster calculation 

Main data structures used in R:

-   **Vector:** A list of items of the same data type, like a series of numbers (numeric) or words (characters).

-   **Matrix:** Tables of numbers with rows and columns, where all elements are the same type.

-   **List:** Collections that can hold different types of data all together, such as numbers, words, or even other lists.

-   **Data Frame:** Like a spreadsheet where each column can have different types of data, useful for tabular data

## Vector

:::{.columns}

:::{.column}
- Basic data structure in R. 
- Store elements in the **same** data type.
- Check the data type by using `typeof()` function
- Check the length of the vector by `length()` function.
- Coercion will be performed if different data type occurs
:::

:::{.column}

```{r}
#| echo: true
#| eval: true
# vector of numerics
# c combines numbers in the brackets to a vector
age <- c(16, 20, 15, 17, 18) 
age
typeof (age)
length(age)
```

:::

:::

See more introduction [here](https://www.datamentor.io/r-programming/vector/).


There are also different ways to automatically generate vectors


## Other examples

There are also different ways to automatically generate vectors

```{r}
#| echo: true
#| eval: true
#| output-location: column
# vectors can be created using existing variables
jason <- 1.69
ray <- 1.78
yuanhua <- 1.72

heights <- c(jason, ray, yuanhua)
# you can give names to items in a vector
names(heights) <- c("Jason","Ray","Yuanhua")
heights
typeof(heights)
```
```{r}
#| echo: true
#| eval: true
#| output-location: column
# rep repeats the first number
# n times based on the second value
repeats <- rep(3, 5)  
repeats
typeof(repeats)
```

```{r}
#| echo: true
#| eval: true
#| output-location: column
# generates a vector of integers 
# from 1 to 12 (inclusive)
series <- 1:12 
series
typeof(series)
```
```{r}
#| echo: true
#| eval: true
#| output-location: column
# generate a vector of random integers by sampling
# 6 values between 1 and 49 (六合彩)
mark6 <- sample(1:49,6)
mark6
typeof(mark6)
```

## Combining vectors

Vectors can also be combined from other vectors.

```{r}
#| echo: true
#| eval: true
# vectors can be combined
new_vector <- c(series,mark6,rep(50,3))
new_vector
typeof(new_vector)
```

Did you notice that `new_vector` is now of type `double` ?

You can spend a few minutes to create vectors and explore what happens to the data type of the vector when you combine them with others.

```{r}
#| echo: true
#| eval: true
# write your code here
myVector <- c(TRUE,2:3,4.0)
myVector
typeof(myVector)
```

## Index

Items in a vector can be directly retrieved by their index

```{r}
#| echo: true
#| eval: true
# Example of using index to retrieve items in a vector
scores <- c(44,56,63,75,77,78,81,90,93,99)
scores
typeof(scores)

scores[5]              # retrieve score at index 5 (i.e. 5th position) 
scores[2:5]            # retrieve score from index 2 to 5
scores[c(2,5,7)]       # retrieve scores from index 2, 5 and 7
scores[-5]             # retrieve all scores except index 5
scores[c(TRUE,FALSE)]  # retrieve scores in odd positions
```

## Exercise

Can you write the code to retrieve the scores from:
```{r}
#| echo: true
#| eval: true
scores <- c(44,56,63,75,77,78,81,90,93,99)
```

1.  the first position

```{r}
#| echo: true
#| eval: true
# write the code to retrieve the 1st position
scores[1]
```

2.  even positions

```{r}
#| echo: true
#| eval: true
# write the code to retrieve even positions
scores[c(FALSE,TRUE)]
```

3.  every 3rd position
```{r}
#| echo: true
#| eval: true
# write the code to retrieve every 3rd position
scores[c(FALSE,FALSE,TRUE)]
```

## Modifying a vector

Using the index it is possible to modify specific items in a vector

```{r}
#| echo: true
#| eval: true
#| output-location: column
# example modifying items in a vector
scores <- c(44,56,63,75,77,78,81,90,93,99)
scores
```
```{r}
#| echo: true
#| eval: true
#| output-location: column
# modify position 5
scores[5] <- 50
scores
```
```{r}
#| echo: true
#| eval: true
#| output-location: column
# modify positions 1 to 4
scores[1:4] <- c(30,40,50,60)
scores
```
```{r}
#| echo: true
#| eval: true
#| output-location: column
# modify position 1 and changes vector to character
scores[1] <- "Twenty"
scores
```

## Matrix

- Two-dimensional data structure. 
- You can think of 2D version of vector
- It has rows and columns, both of which can also have names. 
- To check the dimensions, you can use the `dim()` function.

See more introduction [here](https://www.datamentor.io/r-programming/matrix/).

```{r}
#| echo: true
#| eval: true
myMatrix <- matrix(1:12, nrow=3)
colnames(myMatrix) <- c("C1","C2","C3","C4")    # assign column names
rownames(myMatrix) <- c("R1","R2","R3")         # assign row names
myMatrix
dim(myMatrix)                                   # prints dimension of the matrix 
```

## Indexing a matrix

Accessing items in a matrix is similar to vector but in 2 dimensions

```{r}
#| echo: true
#| eval: true
myMatrix[1, 2]                          # retrieve item in row 1 column 2
myMatrix["R2", "C2"]                    # retrieve item by row and col name in R2, C2
myMatrix[1, 1:2]                        # retrieve items in row 1 and col 1 and 2. 
myMatrix[1:2, c(2, 3)]                  # retrieve items in row 1-2 and col 2-3
```

## Modify values

Modifying items in a matrix is similar to vector but in 2 dimensions. Can you write the code to:

1.  replace row 3, column 3 with the value 7?

```{r}
#| echo: true
#| eval: true
#| output-location: column
myMatrix[3,3] <- 7
myMatrix
```

2.  replace row 1, column 4 using the row and col name with the value -5?

```{r}
#| echo: true
#| eval: true
#| output-location: column
myMatrix["R1","C4"] <- -5
myMatrix
```

3.  delete row 2

```{r}
#| echo: true
#| eval: true
#| output-location: column
myMatrix <- myMatrix[-2, ]
myMatrix
```

## List

- Capable to hold mixed data types.  A `list` can contain a list of any data structure: value, vector, matrix, list, etc.


```{r}
#| echo: true
#| eval: true
#| output-location: column
myList <- list(1.70, TRUE, 1:3, "Ray",list(8,8,4))
str(myList)
```

We can use `str()` function to view the structure of a list (or any object).
```{r}
#| echo: true
#| eval: true
names(myList) <- c("Height","Male","Workdays","Name","hour")    # give names to elements in list
str(myList)
```

## Indexing list

- **double-layer square brackets**, either by numeric index or name. 
- `$` symbol with the name.

```{r}
#| echo: true
#| eval: true
#| output-location: column
# example of retrieving items from a list
myList[[4]] # using index
```

```{r}
#| echo: true
#| eval: true
#| output-location: column
myList[["Name"]]  # using index of Name
```

```{r}
#| echo: true
#| eval: true
#| output-location: column
myList$Name # using the Name key directly
```

Can you retrieve the Height and Name from myList?

```{r}
#| echo: true
#| eval: true
#| output-location: column
myList[c(1,4)] 
```

note that unlike the above, the retrieved values are still in a list as opposed to individual values like what you get with double brackets [[]]
```{r}
#| echo: true
#| eval: true
#| output-location: column
myList[c("Height","Name")]
```

## Converting to/from vector

It is possible to inter-convert list to and from vectors.

```{r}
#| echo: true
#| eval: true
# example of interconverting list to vector and back
myList <- list(1.70, TRUE, 1:3, "Jason")
str(myList)
myVector <- unlist(myList)
myVector
myList_cov <- as.list(myVector)
str(myList_cov)
```

## Data Frame

Data frame is widely used for rectangular data, where each column has the same data type (vector) but different columns can have different data types (like Excel)

The data frame is in fact a special type of `list`: A list of vectors with the same length.

```{r}
#| echo: true
#| eval: true
myDF <- data.frame("Height" = c(1.70, 1.80, 1.68, 1.72), 
                  "Weight" = c(67, 60, 55, 58), 
                  "Name" = c("Jason", "Ray", "Dora", "Yuanhua"), 
                  "Gender" = factor(c("male","male","female","male")))
myDF
myDF$Name[3]          # print name at index 3
levels(myDF$Gender)   # print levels of gender (which is coded as a factor)
```