## Why wrangling matters 

Raw experiment tables are almost never analysis-ready.

- Columns may be wide, messy, or split across files
- We need a clean table: one row per observation, clear columns for predictors/outcomes

**`dplyr`** is how we reshape and filter those tables so plots and models can run.

Keep a before/after picture in mind: messy spreadsheet → tidy analysis table.

## Transforming data with dplyr

Using `tidyverse`  an “umbrella-package” that installs several packages useful for data analysis which work together well such as `tidyr`, `dplyr`, `ggplot2`, `tibble`, etc. 

```{r}
#| echo: true
#| eval: true
# Title: Load tidyverse packages
# tidyverse loads multiple data science packages like ggplot2, dplyr, and tidyr.
library(tidyverse)
```

In this class, we actually use `dplyr` is a package for making tabular data manipulation easier. You can choose using 

```{r}
#| echo: true
#| eval: true
# Title: Load dplyr package
# dplyr provides functions for easy data frame manipulation.
library(dplyr)
```

## Example data - Yeast gene expression

- **Rows** = one gene measured in one nutrient-limitation condition
- **Columns** = gene name, limiting nutrient, growth rate, expression
- **Numbers** = log₂ expression relative to a reference RNA (how strongly that gene was “on”)

We will practice `dplyr` verbs on this table. 

- Source: Brauer et al. (2008), *Coordination of Growth Rate, Cell Cycle, Stress Response, and Metabolic Activity in Yeast*.
- Goal: How does yeast gene expression change with growth rate when a single nutrient is limited?
  
---

- Organism: budding yeast (*Saccharomyces cerevisiae*)
- Exposure: rich media except one limiting nutrient — Glucose, Leucine, Ammonia, Sulfate, Phosphate, or Uracil
- Observations: microarray expression at six growth rates (0.05 to 0.3 h⁻¹)
- Running example: **`LEU1`** (leucine biosynthesis) is turned **up** when leucine is starved, then falls as the restriction is relaxed.

- `brauer2007_tidy.csv`: teaching subset of the  expression table (one row per gene × nutrient × growth rate)

**Reading data**
```{r}
#| echo: true
#| eval: true
# Load the yeast expression table and gene annotation from CSV files for later use.

ydat <- read_csv("https://bdsr.stephenturner.us/data/brauer2007_tidy.csv")
```

## A reminder: Variable assignment

- Only I use `<-` to assign or update the value of a variable, then the variable will be updated to the new value.
- Othersies you just do a calculation and the variable will not be updated.

```{r}
#| echo: true
#| eval: false
# Assign the value '50' to the weight object.
weight <- 50
# Print out weight to the screen (50)
weight
# What's the value of weight plus 10?
weight + 10
# Weight is still 50
weight
# Weight is only modified if we *reassign* weight to the modified value
weight <- weight+10
# Weight is now 60
weight
```

- For rest of the lecture, I won't update the `ydat` object, but instead will show the output of each operation. You can choose to update it if you want to keep the changes.

## Inspecting the data

Getting the first 3 rows of the data frame by `head()`
```{r}
#| echo: true
#| eval: true
# Show the first 3 rows of the expression data.
head(ydat, 3)
```

Getting the summary of the data frame by `summary()`
```{r}
#| echo: true
#| eval: true
# Show a summary of the expression data.
summary(ydat)
```

---

Inspecting the data frame structure by `str()`
```{r}
#| echo: true
#| eval: true
# Show the structure of the expression data.
str(ydat)
``` 

## Pipeline mindset

Without `dplyr`, if we want to manipulate data, we can still use the following vanilla way to perform the task :


```{r}
#| echo: true
#| eval: false
# Title: Base R data manipulation example
# Show how row selection, counting, and column updates work without dplyr.

# Select rows 2 to 5 from the expression table.
ydat <- ydat[2:5,]

# Count how many Leucine-limited and Glucose-limited rows are in the selection.
sum(ydat$nutrient == 'Leucine')
sum(ydat$nutrient == 'Glucose')

# Overwrite the nutrient column with a logical value indicating leucine limitation.
ydat$nutrient = ydat$nutrient == 'Leucine'

```


::: {.callout-note}
## Discussion question

- What are the common features for these operations
- What are the pitfalls in using this approach ?
:::

## Pipe operator |>

To prevent us from keep doing like a matryoshka doll

```{r}
#| echo: true
#| eval: false
# Title: Nested function call example
# This shows the same work as a pipeline, but in nested form.
h(g(f(data_frame)))

```

R by default supports Pipe operator |> We can rewrite the above code as 

:::{.columns}
:::{.column}

```{r}
#| echo: true
#| eval: false
# Pipe operator example
# Use the pipe operator to write the same 
# steps in a clearer, left-to-right order.
data_frame |> 
  f() |> 
  g() |> 
  h()
```

:::

:::{.column}
```{r}
#| echo: true
#| eval: false
# This is the equivalent code of the left one.

data_frame <-f(data_frame)
data_frame <-g(data_frame)
data_frame <-h(data_frame)

```
:::
:::

'|>' takes the output from the left hand side and then **pipes** it to be the first argument of the right function.

::: {.callout-note}
## Discussion question

- What are the benefits of using pipe?
  
:::

## Select columns or rows

- Select the designated columns by `select(COLUMN_NAMES)`


```{r}
#| echo: true
#| eval: true
# Title: Select expression columns
ydat |>
    select(symbol, nutrient, rate, expression) |> # Keep gene, nutrient, growth rate, and expression.
    head(3) # This line is for display control, you can omit it and the previous |>.
```

## Select by conditions

- Select the designated rows by `filter(CONDITION)`

```{r}
#| echo: true
#| eval: true
# Title: Filter rows by gene
ydat|>
    filter(symbol == 'LEU1') |> # Keep only the leucine biosynthesis gene LEU1.
    head(3) # This line is for display control, you can omit it and the previous |>.
```

::: {.callout-warning}
## Common error

`filter()` function has different meaning in other R default library. If the error message occurs, you can try replaced it by `dplyr::filter()` instead
:::

## Manipulating columns

- You can create new column by `mutate()`
```{r}
#| echo: true
#| eval: true
# Title: Create or modify columns
ydat |>
    mutate( # expression is log2(sample/reference); 2^expression recovers the raw ratio.
              signal = 2^expression,
              restriction  = "Nutrient-limited"
              ) |>
    head(3)
```

## Perform summary statistics

Frequently we will need to provide summary statistics for your data, we can use `summarize()` to do it


```{r}
#| echo: true
#| eval: true
# Title: Compute summary statistics

ydat |>
    summarize( # Summarize expression with count, mean, standard deviation, and LEU1 row count.
              n=n(),
              mean_expr=mean(expression),
              sd_expr=sd(expression),
              n_leu1=sum(symbol=='LEU1')
    )
```


Some common summary function could be

:::{.columns}

:::{.column }

- `mean()`
- `sd()`
- `max()`, `min()`

:::

:::{.column }

- `sum()`
- `n()`

:::

:::
## Summary statistics by group

:::{.columns}

:::{.column width=33%}

- `group_by()` and `summarize()` provide summary statistics by group.

:::

:::{.column width=65%}
```{r}
#| echo: true
#| eval: true
# Title: Grouped summary by nutrient
ydat |>
# Group the expression table by limiting nutrient,
    group_by(nutrient) |> 
#  and compute summary statistics for each group.
    summarize(      
              n=n(),
              mean_expr=mean(expression)
                  )
```

:::

:::

## Condition matching

:::{.columns}

:::{.column width=33%}

`if_else(CONDITION, 'TRUE_VALUE', 'FALSE_VALUE')`  for dichotomous processing

:::

:::{.column width=65%}


```{r}
#| echo: true
#| eval: true
# Title: Create binary restriction status
# Use if_else() to add a restriction label based on growth rate.
ydat |>
    select(symbol, nutrient, rate, expression) |>
    mutate( restriction = if_else(
        rate == 0.05, "Severe restriction", "Relaxed"
    )) |>
    head(3)
```
:::

:::

## Condition matching

:::{.columns}

:::{.column width=33%}

- `case_when(Condition ~~ "Label")` to handle sequential logical control

:::

:::{.column width=65%}
```{r}
#| echo: true
#| eval: true
# Title: Create categorical restriction stage
# Use case_when() to assign descriptive labels for selected growth rates.
ydat |>
    select(symbol, nutrient, rate, expression) |>
    mutate( restriction = case_when(
        rate == 0.05 ~ "Severe restriction",
        rate == 0.15 ~ "Moderate restriction",
        rate == 0.3 ~ "Relaxed restriction"
    ))|>
    head(3)
```
:::

:::

## Best practice of data wrangling

1. The "Read-Write-Verify" Loop

   - Write a step
   - Run it immediately
   - Verify the output

    **DO NOT** propagate error and raise your hand

2. Look up documentations
3. Make your code **readable**
4. Trial and error

# Advance topics

## Renaming dataframe columns

- You can rename column by `rename()`
```{r}
#| echo: true
#| eval: true
# Title: Rename columns
ydat |> # Rename selected expression columns to make them more readable.
    rename(
              Gene = symbol,
              Nutrient = nutrient,
              Rate = rate
              ) |>
    head(3)
```

## Groupby multiple conditions

You can group by multiple groups together, but note that they need to be stated in the same `group_by()` brackets

```{r}
#| echo: true
#| eval: false
# Title: Multi-group summary
ydat |>
    group_by(nutrient, rate) |> # Group by both limiting nutrient and growth rate,
    summarize(                   #  then summarize the combined groups.
              n=n(),
              mean_expr=mean(expression),
              sd_expr=sd(expression),
    )
```
<!-- 
## Combining Dataframes

It’s rare that a data analysis involves only a single table of data. 

Consider we have data_frame `brauer2007_tidy.csv` and `brauer2007_sysname2go.csv`

`brauer2007_sysname2go.csv` links the systematic name to gene ontology information, let's call it `annot`
```{r}
annot <- read_csv("https://bdsr.stephenturner.us/data/brauer2007_sysname2go.csv")
```

##
There are four options  

- `inner_join(x, y)`: Keeps rows present in both tables.
- `left_join(x, y)`: Keeps all rows from x, matching rows from y.
- `right_join(x, y)`: Keeps all rows from y, matching rows from x.
- `full_join(x, y)`: Keeps all rows from both tables.

## Combining Dataframes

Join expression rows to Gene Ontology labels by the yeast systematic name.You can combine the expression table and the annotation table together. 

```{r}
#| echo: true
#| eval: true
# Title: Join expression with gene annotation
# Combine the yeast expression table with GO annotation by systematic gene ID.
# ydat |> inner_join(annot, by = 'systematic_name') |>
#     select(symbol, nutrient, expression, bp) |>
#     head(3)
```
 -->
