
## Example data - Gene Expression

- Source: Blackmore et al. (2017), *The effect of upper-respiratory infection on transcriptomic changes in the CNS*. 
- Goal: Determine the effect of an upper-respiratory infection on changes in RNA transcription occuring in the cerebellum and spinal cord post infection
- Experiment sample: Gender matched eight week old C57BL/6 mice  
- Exposure: With or without Influenza A infection
- Observations: RNA-seq count of genes in the cerebellum and spinal cord tissues
- Observation time: days 0 (non-infected), 4 and 8.

## Task - Read two dataframe

Please read the data to your RStudio for further analysis

:::{.hide}

```{r}
#| echo: false
#| eval: true
# Title: Load example datasets silently
# Load the rna count matrix and sample metadata from CSV files for later use.

rnaseq <- read.csv("../../../assets/data/dplyr/count_matrix.csv")
meta <- read.csv("../../../assets/data/dplyr/sample_metadata.csv")
```

:::

```{r}
#| echo: true
#| eval: true
# Title: Read files and inspect data
# Read the count matrix and metadata, then preview the first few rows.
# --- Your code starts here ---

# rnaseq <- read.csv("WRITE DOWN YOUR PATH HERE")
# meta <- read.csv("WRITE DOWN YOUR PATH HERE")

# --- Your code ends here ---

# Show the first 3 rows and first 6 sample columns of the expression data.
head(rnaseq[, 1:6], 3)

head(meta, 3)
```

::: {.callout-note}
## Discussion question

- What are the pros and cons using absolute path and relative path ?
- What will be the best practce to put the data file ?
:::

## Transforming data with dplyr

`dplyr` is a package for making tabular data manipulation easier. You can choose using 

```{r}
#| echo: true
#| eval: true
# Title: Load dplyr package
# dplyr provides functions for easy data frame manipulation.
library(dplyr)
```

or 

using `tidyverse`  an “umbrella-package” that installs several packages useful for data analysis which work together well such as `tidyr`, `dplyr`, `ggplot2`, `tibble`, etc. 

```{r}
#| echo: true
#| eval: true
# Title: Load tidyverse packages
# tidyverse loads multiple data science packages like ggplot2, dplyr, and tidyr.
library(tidyverse)
```

## Pipeline mindset

Without `dplyr`, if we want to manipulate data, we can still using 0the following vanila way to perform the task :


```{r}
#| echo: true
#| eval: false
# Title: Base R data manipulation example
# Show how row selection, counting, and column updates work without dplyr.

# Select rows 2 to 5 from the metadata.
meta <- meta[2:5,]

# Count how many female and male mice are in the selected rows.
sum(meta$sex == 'Female')
sum(meta$sex == 'Male')

# Overwrite the sex column with a logical value indicating male samples.
meta$sex = meta$sex == 'Male'

```


::: {.callout-note}
## Discussion question

- What are the common features for these operations
- What are the pitfalls in using this approach ?
:::

## Pipe operator `%>%`

To prevent us from keep doing like a matryoshka doll

```{r}
#| echo: true
#| eval: false
# Title: Nested function call example
# This shows the same work as a pipeline, but in nested form.
h(g(f(data_frame)))

```

Libraries under `tidyverse`  support Pipe operator `%>%`

We can rewrite the above code as 

```{r}
#| echo: true
#| eval: false
# Title: Pipe operator example
# Use the pipe operator to write the same steps in a clearer, left-to-right order.
data_frame %>% 
  f() %>% 
  g() %>% 
  h()
```

`%>%` takes the output from the left hand side and then **pipes** it to be the first argument of the right function.

::: {.callout-note}
## Discussion question

- What are the benefits of using pipe?
  
:::

## Select columns or rows

- Select the designated columns by `select(COLUMN_NAMES)`


```{r}
#| echo: true
#| eval: true
# Title: Select metadata columns
meta %>%
    select(age, sex, infection) %>% # Keep only the age, sex, and infection columns.
    head(3) # This line is for display control, you can omit it and the previous %>%.
```

- Select the designated rows by `filter(CONDITION)`

```{r}
#| echo: true
#| eval: true
# Title: Filter rows by sex
meta %>%
    filter(sex == 'Female') %>% # Keep only female samples.
    head(3) # This line is for display control, you can omit it and the previous %>%.
```

::: {.callout-warning}
## Common error

`filter()` function has different meaning in other R default library. If the error message occurs, you can try replaced it by `dplyr::filter()` instead
:::

## Manipulating columns

- You can rename column by `rename()`
```{r}
#| echo: true
#| eval: true
# Title: Rename columns
meta %>% # Rename selected metadata columns to make them more readable.
    rename(
              Sex = sex,
              Time = time,
              Infection = infection
              ) %>%
    head(3)
```

- You can create new column by `mutate()`
```{r}
#| echo: true
#| eval: true
# Title: Create or modify columns

meta %>%
    mutate( # Use mutate() to update the time column and add a new infection flag.
              time = time + age,
              infection  = "No"
    ) %>%
    head(3)
```

## Perform summary statistics

Frequently we will need to provide summary statistics for your data, we can use `summarize()` to do it


```{r}
#| echo: true
#| eval: true
# Title: Compute summary statistics

meta %>%
    summarize( # Summarize the metadata with count, mean, standard deviation, and female sample count.
              n=n(),
              mean_time=mean(time),
              sd_time=sd(time),
              num_female=sum(sex=='Female')
    )
```


Some common summary function could be

- `mean()`
- `sd()`
- `max()`, `min()`
- `sum()`
- `n()`

## Summary statistics by group

:::{.columns}

:::{.column width=33%}

- `group_by()` and  `summarize`p rovide summary statistics by group.
- You can group by multiple groups together, but noted that they need to be stated in the same `group_by()` brackets

:::

:::{.column width=65%}
```{r}
#| echo: true
#| eval: true
# Title: Grouped summary by infection

meta %>%
    group_by(infection) %>% # Group the metadata by infection status,
    summarize(      #  and compute summary statistics for each group.
              n=n(),
              mean_time=mean(time),
    )
```

```{r}
#| echo: true
#| eval: false
# Title: Multi-group summary
meta %>%
    group_by(infection, sex) %>% # Group by both infection status and sex,
    summarize(                   #  then summarize the combined groups.
              n=n(),
              mean_time=mean(time),
              sd_time=sd(time),
    )
```
:::

:::

## Condition modification

:::{.columns}

:::{.column width=33%}

Using `if_else(CONDITION, TRUE, FALSE)` or `case_when(CONDITION ~ OUTCOME, ...)` to rematch condition into certain value/ ordinal items

- `if_else` is only for dichotomous processing
- `case_when` can handle sequential logical control

:::

:::{.column width=65%}


```{r}
#| echo: true
#| eval: true
# Title: Create binary infection status
# Use if_else() to add a new infection_status column based on time value.
meta %>%
    select(sample, age, sex, time) %>%
    mutate( infection_status = if_else(
        time == 0, "Not infected", "Infected"
    )) %>%
    head(3)
```
```{r}
#| echo: true
#| eval: true
# Title: Create categorical infection stage
# Use case_when() to assign descriptive labels for each time point.
meta %>%
    select(sample, age, sex, time) %>%
    mutate( infection_status = case_when(
        time == 0 ~ "Not infected",
        time == 4 ~ "Early stage",
        time == 8 ~ "Late stage"
    ))%>%
    head(3)
```
:::

:::

## Combining Dataframes

Consider we have data_frame `sample_metadata.csv` and `count_matrix.csv`. If we want to combine them together. We can use `join()`

There are four options  

- `inner_join(x, y)`: Keeps rows present in both tables.
- `left_join(x, y)`: Keeps all rows from x, matching rows from y.
- `right_join(x, y)`: Keeps all rows from y, matching rows from x.
- `full_join(x, y)`: Keeps all rows from both tables.

## Combining Dataframes

First I transpose the RNA count matrix. (The below code chunk is out of the scope, you might ignore why it works)
```{r}
#| echo: true
#| eval: true
# Title: Transform RNA count matrix layout
# Convert the RNA data from wide to long and back to wide format by gene.
rnaseq <- rnaseq %>% 
           pivot_longer(cols=-gene, names_to="sample", values_to = "value") %>%
  pivot_wider(names_from = gene, values_from = value)
```

You can combine meta data and rna sequence data together.


```{r}
#| echo: true
#| eval: true
# Title: Join metadata with RNA data
# Combine the sample metadata with the RNA expression table by sample ID.
meta %>% left_join(rnaseq, by = 'sample') %>% select(1:12) %>% head(3)
```

## Best practice of data wrangling

1. The "Read-Write-Verify" Loop

   - Write a step
   - Run it immediately
   - Verify the output

    **DO NOT** propogate error and raise your hand

2. Look up documentations
3. Make your code **readable**
4. Try and error

