## Reading and writing data to file

For reading and writing data we recoomend use the library under `tidyverse`

```{r}
#| echo: true
#| eval: false
#install tidyverse (package with lots of functions to handle data in R)
if (!requireNamespace("tidyverse", quietly = TRUE)) {    # checks if already installed
  install.packages("tidyverse")
}
library(tidyverse)                                       # load the package
```

:::{.aside}
To be specific it is `readr` library. But `tidyverse` can install all the library you need for this course
:::

## Loading/saving data from folders

- By default, R sets the working directory to your `Documents` folder.
- Use `setwd(PATH/TO/NEW/WORKING/DIRECTORY)` to change the working directory.
- Use `getwd()` to see your current working directory.

Consider the following file structure

```md
/ (root)
├── home
│   └── user
│       ├── documents
│       │   ├── course
│       │   │   ├── THIS_RMD.rmd
│       │   │   └── myData.csv
│       │   └── notes.txt
│       └── downloads
|           └── IMAGE.png
└── var
```

## Read the data

Using `read_csv()` to read the [data](https://github.com/StatBiomed/BMDS-book/raw/refs/heads/main/notebooks/chapter1-R/myData.csv)

```{r}
#| echo: true
#| eval: false

myData <- read_csv('myData.csv')
```

- `read_csv()` need to specify WHERE do you put the file by **Absolute path** or **Relative path**
- If you are using Rmarkdown (`.rmd`) to run the above code then you need to put in the same directory
- Otherwise you need to put at your current working directory

## File path

:::{.columns}

:::{.column width=40%}

```md
/ (root)
├── home
│   └── user
│       ├── documents
│       │   ├── course
│       │   │   ├── THIS_RMD.rmd
│       │   │   └── myData.csv
│       │   └── notes.txt
│       └── downloads
|           └── IMAGE.png
└── var
```

:::


:::{.column width=57%}

There are two ways of specifying a path:

1. **Absolute path**
   The full location of the file or directory.
2. **Relative path**
   The location of the file or directory relative to the current directory.

:::

:::


##  Absolute Path

:::{.columns}

:::{.column width=40%}

```md
/ (root)
├── home
│   └── user
│       ├── documents
│       │   ├── course
│       │   │   ├── THIS_RMD.rmd
│       │   │   └── myData.csv
│       │   └── notes.txt
│       └── downloads
|           └── IMAGE.png
└── var
```

:::


:::{.column width=57%}
**Absolute Path**

An absolute path starts from the root directory (`/`) and specifies the full path to a file or directory. 

- Path to `THIS_RMD.rmd`: 
  `/home/user/documents/course/THIS_RMD.rmd`
- Path to `IMAGE.png`:
  `/home/user/downloads/IMAGE.png`

:::

:::

:::{.callout-tip}
## Reminder for Windows system

- You will find your file path using `\` instead of `/` in your file explorer. In R, you need to transform these `\`s to `/` if you directly copy and paste the file path.
- You need to indicate your drive, so it usually looks like `C:\Users\Ray\Blah\Blah`
::: 



##  Relative path

:::{.columns}

:::{.column width=40%}

```md
/ (root)
├── home
│   └── user
│       ├── documents
│       │   ├── course
│       │   │   ├── THIS_RMD.rmd
│       │   │   └── myData.csv
│       │   └── notes.txt
│       └── downloads
|           └── IMAGE.png
└── var
```
:::


:::{.column width=57%}

**Relative Path**

A relative path is defined relative to the current working directory. 

- `getwd()` to get current working directory
- `setwd('A PATH')` to set new working directory

- `./` refers to the current directory,
- `../` means "go up one directory level".

:::

:::

Assume that we are working on `THIS_RMD.rmd`, so the current directory is `/home/user/documents/course`.

- Path to `myData.csv` : `./myData.csv` or `myData.csv`
- Path to `IMAGE.png`: `../../downloads/IMAGE.png`
