# Fuoundation of R

## Introduction to R

R is a programming language popular for statistical computing, elegant graphics, and genomic data analysis.

**Features** 

- Free and open-source software
- Easy to get started for scientific computing

**Slide conventions**

- Use code formatting for special terms and examples, such as `This` or **`This are code`**.

```{r}
#| echo: true
#| eval: true
# This is a code chunk
```

## Arithmetic operations

R provides many arithmetic and statistical operations like `+`, `-`, `*`, `/`, `%%`, and `^`

```{r}
#| echo: true
#| eval: true
# example of basic operation
2+2
5*3
10/3
10%%3 # %% returns the remainder of the division
2^3 # ^ is the power operation
```


## Variables

Variables store information. To save a value, assign it to a variable. The assignment operator in R is `<-`. Occasionally, you may also see `=` used for assignment, but `<-` is the preferred style in R.

$$
\small
\text{Variable <- Value}
$$

Benefits of using variables:

- Store and manipulate data in a flexible way.
- Make code more readable and maintainable.
- Allow dynamic programming and automation.
```{r}
#| echo: true
#| eval: true

# example of assigning variable called age

age <- 18
age

name <- "John"
name
```

## Naming conventions for variables

Variables can be called just about anything, but there are some rules in R:

-   Special characters like **`$`**, **`#`**, **`%`**, or spaces are **not allowed**.
-   A variable name must **start with a letter or a dot (.)**, but if it starts with a dot, the next character cannot be a digit.
-   Variable names are **case-sensitive** (e.g., **`age`**, **`Age`**, and **`AGE`** are different variables).
-   You cannot use particular keywords such as **`TRUE`**, **`FALSE`**, **`if`**, **`else`**, or **`function`** as variable names as they are reserved for other purposes in R.


## Data types

Variables can store different types of data, and R has several built-in data types to support them. The data type of a variable determines what kind of values it can hold and what operations can be performed on it.

You can use the `class()` and `typeof()` functions to check the class and data type of any variable. R has five common data types:

1. Numeric
2. Integer
3. Complex
4. Logical
5. Character

## Numeric (double)

The `numeric` data type is for numbers.

- Can be whole numbers or decimals.
- The **default** numeric type in R.
- Stores values in double precision, so `typeof()` reports `double`.

```{r}
#| echo: true
#| eval: true
# example of numeric type

height <- 1.70
height
class(height)
typeof(height)
```

## Integer

The `integer` data type stores whole numbers.

- Use the capital `L` suffix to create an integer value.
- Convert a numeric value to integer using `as.integer()`.
- Convert an integer to numeric using `as.numeric()`.
```{r}
#| echo: true
#| eval: true
# example of integer type

siblings <- 2L
siblings 
typeof(siblings)
siblings <- as.numeric(siblings)
typeof(siblings)
```

## Logical

In R, the `logical` data type can be:

- `TRUE` (carries a value of 1)
- `FALSE` (carries a value of 0)

*Tips*: Logical values can behave like numbers. Numeric operations are possible on logical values, but use caution.

```{r}
#| echo: true
#| eval: true
# example of logic type

male <- TRUE
male
typeof(male)
male <- male-1
male
typeof(male)
```

## Character

The `character` data type stores text strings, including letters, numbers, and symbols. Strings in R can be written with single or double quotes.

```{r}
#| echo: true
#| eval: true
# example of character type

name <- "Ray"
name
typeof(name)
age <- "25"
age
typeof(age)
```

## Comparison operations

Comparison operators compare two values and return a logical value `TRUE` or `FALSE`.
```{r}
#| echo: true
#| eval: true
# example of basic operations
2 == 2 # Equal
5 < 3 # Less than
10 >= 3 # Greater than or equal to
10 != 3 # Not equal
## Weird behavior !!
1 == "1"
1 == TRUE
"1" == TRUE
```


## Coercion

while performing any operations mixed with different data types in R, the values are automatically converted to the most general data type by **coercion**. The order is as below

$$
\text{logical} < \text{integer} < \text{numeric} < \text{character}
$$


```{r}
#| echo: true
#| eval: true
1.2 + TRUE
# 2.3 + "1" <= This line will result in error
## Weird behavior !!
1 == "1"
1 == TRUE
"1" == TRUE
```

## Functions

Functions are used to perform tasks on input values. For example, the `sin()` function calculates the sine of a number. The input to the function is given inside the parentheses.

:::{.columns}
:::{.column width="49%"}
$$
\sin (\pi) = 0
$$

:::

:::{.column width="49%"}

```{r}
#| echo: true
#| eval: true
# example of sin() function
sin(pi)
```

:::

:::

- `sin` is the name of the function, and `pi` is the input. The output of the function is `0`.
- Why is the output not exactly `0`? This is because π is stored as a double and is accurate to about 15–16 decimal places. As a result, `sin(pi)` is a very small number close to zero, but not exactly zero.

## More function examples

Another example is `floor()`, which rounds a number down to the nearest whole number.


```{r}
#| echo: true
#| eval: true
# example of floor(sin(pi)) function
sin_pi <- sin(pi)
floor(sin_pi)
```


- The output of `floor(sin(pi))` is `0`. This is because the `floor()` function rounds down the value of `sin(pi)` to the nearest whole number.
- R has many predefined functions, and you can also create your own functions. See the official [documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html) for more information.
- Within RStudio, you can use `?` to get help on a function. For example, `?floor` gives information about the `floor()` function.

## Writing your own function

You can customize functions in R to perform specific tasks. Benefits of using functions:

- Make your code more organized and readable.
- Allow for code reusability and modularity.

The format for writing a function in R is:

$$
\begin{aligned}
\textit{function_name} &\leftarrow \text{function}(\textit{arguments}) \{ \\
& \textit{your code} \\
& \}
\end{aligned}
$$

## BMI example

Consider the following example of a function that calculates BMI:

```{r}
#| echo: true
#| eval: true
# First define a function to calculate BMI

# weight in kg, height in meters
calculate_bmi <- function(weight, height) {
  bmi <- weight / (height^2)
  return(bmi)
}
# Example usage:
yuanhua_weight <- 66.2
yuanhua_height <- 1.71
ray_weight <- 68.1
ray_height <- 1.85
calculate_bmi(yuanhua_weight, yuanhua_height)
calculate_bmi(ray_weight, ray_height)

```