
## Overview of linear regression

In linear regression, our fundamental objective is to explain the variability observed in the outcome variable ($Y$) using one or more explanatory variable ($X_1, X_2, \dots, X_k$). 

:::{.hide}
$$
\hat{Y}_i = b_0 + b_1 X_{1i} + b_2 X_{2i} + \dots + b_k X_{ki}
$$
:::

It is crucial to know

- Does $X_1, X_2, \dots, X_k$ really make explaination to $Y$
- How good is the model performance?

## Evaluating linear regression - from an observation


- The regression diagnostics lies in the **Analysis of Variance (ANOVA)**
- We can decompose the error of you prediction as:

:::{.hide}
$$
(Y_i - \hat{Y}_i) = (Y_i - \bar{Y})-(\hat{Y}_i - \bar{Y}) 
$$
:::

Where:

- **$(Y_i - \hat{Y}_i) = e_i$** = Unexplained residual deviation (the error remaining after fitting the model).
- **$(Y_i - \bar{Y})$** = Total deviation of the data point from the baseline benchmark (sample mean).
- **$(\hat{Y}_i - \bar{Y})$** = Explained deviation (the difference between the model's fitted value and the sample mean).

## A summary of error 

Squaring both sides and summing across all $n$ observations yields the fundamental sum-of-squares identity:

:::{.hide}

$$ 
\text{SSE} = \text{SST} - \text{SSR}
$$

:::

Where 

- SST (Total Sum of Squares): $\sum_i (Y_i - \hat{Y}_i)^2$
- SSE (Error Sum of Squares): $\sum_i (Y_i - \bar{Y})^2$
- SSR (Regression Sum of Squares): $\sum_i (\hat{Y}_i - \bar{Y})^2$

   
:::{.aside}
(Note on terminology: In some literature, notably British and econometrics, SSR is termed "Explained Sum of Squares" (ESS) and SSE is termed "Residual Sum of Squares" (RSS). Always check definitions in your specific context).
:::

## Mathematical Formulations & Intuitions


| Metric | Formula | Degrees of Freedom ($df$) | Conceptual Intuition |
| :-- | :-- | :-- | :----- |
| **SST** | $\sum_{i=1}^n (Y_i - \bar{Y})^2$ | $n - 1$ | Total baseline variation in the outcome $Y$ without using any predictors. |
| **SSR** | $\sum_{i=1}^n (\hat{Y}_i - \bar{Y})^2$ | $k$ | Variation successfully accounted for by the regression model. |
| **SSE** | $\sum_{i=1}^n (Y_i - \hat{Y}_i)^2$ <br> $= \sum_{i=1}^n e_i^2$ | $n - k - 1$ | Unexplained variation; sum of squared discrepancies between observed points and regression surface. |

:::{.aside}
- $n$ represents the sample size
- $k$ represents the number of predictors, excluding the intercept.
:::
---

## Goodness-of-Fit: $R^2$ 

$R^2$ (The coefficient of determination) quantifies the proportion of total variation in the response variable explained by the linear combination of predictors:

:::{.hide}

$$
R^2 = \frac{\text{SSR}}{\text{SST}} = 1 - \frac{\text{SSE}}{\text{SST}}
$$
:::

**Interpretation Rules**

- **Range:** In standard OLS with an intercept, $0 \le R^2 \le 1$ (or $0\%$ to $100\%$).
- **Interpretation:** If $R^2 = 0.74$, exactly **74% of the variability in $Y$ is explained by the predictors** included in the model, while the remaining 26% is attributable to residual noise or omitted factors.
- **Limitation:** $R^2$ is non-decreasing with respect to the addition of predictors. Adding completely unrelated random noise variables will force $\text{SSE}$ to stay equal or decrease, artificially inflating $R^2$.

## Adjusted $R^2$ ($R^2_{\text{adj}}$)

The **Adjusted $R^2$** incorporates degrees of freedom to penalize excessive parameterization and prevent overfitting, :

:::{.hide}
$$
R^2_{\text{adj}} = 1 - \left[ \frac{\text{SSE} / (n - k - 1)}{\text{SST} / (n - 1)} \right] = 1 - \left( \frac{\text{MSE}}{\text{MST}} \right)
$$
:::

Alternatively expressed via $R^2$:
$$
R^2_{\text{adj}} = 1 - (1 - R^2) \frac{n - 1}{n - k - 1}
$$

**Key Properties**

- Increases only if a newly added predictor reduces $\text{SSE}$ by a margin greater than what would be expected by pure chance.
- Can occasionally yield a negative value when model fit is negligible; in such cases, it is effectively interpreted as zero.


## $F$-Test for Model Significance

 $F$-Test** evaluates whether the collection of predictors collectively explains a statistically significant amount of variation.

**Hypotheses**

- **Null Hypothesis ($H_0$):** All slope coefficients are simultaneously zero.
  
:::{.hide}

$$
H_0: \beta_1 = \beta_2 = \dots = \beta_k = 0
$$

:::

- **Alternative Hypothesis ($H_1$):** At least one slope coefficient is non-zero.
  
:::{.hide}

$$
H_1: \text{There is at least one} \beta_j \neq 0
$$

:::

**Difference on individual testing on $\beta_i$**

- $\beta_i$ is a $t$-test
- Repeated testing $\beta_i$  cause p-value inflation

## Mean Squares: Variance Between vs. Variance Within

Before calculating the $F$-statistic, we normalize the sums of squares by their respective degrees of freedom:

**Mean Square Regression (MSR):**

:::{.hide}

$$
\text{MSR} = \frac{\text{SSR}}{k}
$$

:::
**Mean Square Error (MSE):**

:::{.hide}

$$
\text{MSE} = \frac{\text{SSE}}{n - k - 1} = s_e^2 \quad (\text{an unbiased estimator of } \sigma^2)
$$

:::

## Constructing the $F$-Statistic

The $F$-statistic is the ratio of explained variance per degree of freedom to unexplained variance per degree of freedom:

:::{.hide}
$$
F = \frac{\text{MSR}}{\text{MSE}} = \frac{\text{SSR} / k}{\text{SSE} / (n - k - 1)}
$$
:::

It can also be written directly in terms of $R^2$:

$$
F = \frac{R^2 / k}{(1 - R^2) / (n - k - 1)}
$$

## Sampling Distribution & Decision Rule:

Under the null hypothesis $H_0$, and assuming **normally distributed homoscedastic** errors:
$$
F \sim F(df_1 = k, \, df_2 = n - k - 1)
$$

- **Decision:** Reject $H_0$ at significance level $\alpha$ if $F > F_{\text{crit}}(\alpha, k, n - k - 1)$ or if $p\text{-value} < \alpha$.
- **Conclusion:** Rejecting $H_0$ confirms that the regression model explains significantly more variance than a naive "intercept-only" model (sample mean $\bar{Y}$).


## 5. The Standard ANOVA Table for Regression

Statistical software (e.g., R, Python `statsmodels`, Stata, SAS) formats these calculations into an ANOVA table:

| Source of Variation | Sum of Squares ($SS$) | Degrees of Freedom ($df$) | Mean Square ($MS$) | $F$-Statistic | $p$-value |
| :--- | :-- | :--- | :--- | :--- | :--- |
| **Regression (Model)** | $\text{SSR}$ | $k$ | $\text{MSR} = \frac{\text{SSR}}{k}$ | $F = \frac{\text{MSR}}{\text{MSE}}$ | $P(F_{(k, n-k-1)} \ge F)$ |
| **Residual (Error)** | $\text{SSE}$ | $n - k - 1$ | $\text{MSE} = \frac{\text{SSE}}{n - k - 1}$ | — | — |
| **Total** | $\text{SST}$ | $n - 1$ | $\text{MST} = \frac{\text{SST}}{n - 1}$ | — | — |

---

## Annotated Worked Example

:::{.columns}

:::{.column width=45%}

Research purpose: Factors affecting infant birth weight ($n = 189$).

- Outcome ($Y$): bwt (Infant birth weight in grams).


:::


:::{.column width=52%}

- Predictors ($X$): Maternal age (age), 
  - maternal weight at last menstrual period (lwt),
  -  smoking status during pregnancy (smoke), 
  -  history of hypertension (ht), 
  -  uterine irritability (ui)

:::

:::

```{r}
#| echo: true
#| eval: false

# Load MASS (pre-installed in R)
library(MASS)
data(birthwt)

# Convert binary indicator variables into explicit factors
birthwt$smoke <- factor(birthwt$smoke, labels = c("Non-Smoker", "Smoker"))
birthwt$ht    <- factor(birthwt$ht, labels = c("No", "Yes"))
birthwt$ui    <- factor(birthwt$ui, labels = c("No", "Yes"))
```

## Regression results

```{r}
#| echo: true
#| eval: false

model <- lm(bwt ~ age + lwt + smoke + ht + ui, data = birthwt)
summary(model)
```

```text
Residuals:
     Min       1Q   Median       3Q      Max 
-1675.13  -464.38    -1.02   459.70  1779.69 

Coefficients:
            Estimate Std. Error t value Pr(>|t|)    
(Intercept) 2506.354    291.759   8.591 3.73e-15 ***
age            3.648      9.432   0.387  0.69935    
lwt            4.388      1.691   2.594  0.01024 *  
smokeSmoker -240.848    100.413  -2.399  0.01746 *  
htYes       -643.958    207.232  -3.107  0.00219 ** 
uiYes       -547.068    139.937  -3.909  0.00013 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 671.4 on 183 degrees of freedom
Multiple R-squared:  0.1747,	Adjusted R-squared:  0.1522 
F-statistic:  7.75 on 5 and 183 DF,  p-value: 1.234e-06
```

## ANOVA table

```{r}
#| echo: true
#| eval: false

library(olsrr)
olsaov<-ols_regress(model)$anova
olsaov$anova
```


```text
                                 ANOVA                                   
------------------------------------------------------------------------
                    Sum of                                              
                   Squares         DF    Mean Square      F        Sig. 
------------------------------------------------------------------------
Regression    17469477.719          5    3493895.544     7.75    0.0000 
Residual      82500178.091        183     450820.645                    
Total         99969655.810        188                                   
------------------------------------------------------------------------
```

## Key Metrics Calculations

1. **$R^2$ Calculation:**
   $$R^2 = \frac{17469477.719 }{ 99969655.810} = \quad (17.4\%)$$
   *Interpretation:* 12.6% of the variance in biomarker reduction is explained by maternal weight at last menstrual period, baseline metabolic index, and age.

2. **Adjusted $R^2$ Calculation:**
   $$R^2_{\text{adj}} = 1 - \left( \frac{450820.645}{99969655.810/ 188} \right) = = 1 - 0.8478 = 0.1522 $$
   *Interpretation:* After penalizing for adding three model parameters, the model accounts for approximately 70.2% of the variance.

3. **$F$-Test Evaluation:**
   $F_{(5, 183)} = 7.75, \quad p < 0.0001$$
   *Conclusion:* Since the $p$-value is well below $\alpha = 0.05$, we reject $H_0$. The model as a whole explains a statistically significant amount of variation compared to a flat horizontal line.



## Simple Linear Regression: $t$-test vs. $F$-test

In simple linear regression ($k = 1$):
$$
F_{(1, n-2)} = (t_{\beta_1})^2
$$

The two-sided $t$-test on the single slope coefficient $\beta_1$ and the overall $F$-test are mathematically identical and yield the exact same $p$-value.

## High $F$-statistic, but no individual significant $t$-statistics?

This classic paradox typically occurs due to **severe multicollinearity**. 

- The predictors are highly correlated with each other.
- The overall $F$-test confirms that the set of predictors collectively accounts for substantial variation.
- However, the variance inflation factor (VIF) is high, inflating individual standard errors and making it impossible for the individual $t$-tests to isolate which specific variable is doing the work.

## Low $R^2$ with a Statistically Significant $F$-test

- **Common in noisy biological or social science data** (e.g., $R^2 = 0.08$, $p < 0.001$ with $n = 5,000$).
- This means the effect is genuine, real, and reproducible (not a sampling artifact), but individual-level variation remains high due to unmeasured factors.

## Pitfalls to Avoid

1. **Confusing $R^2_{\text{adj}}$ with Causality:** A high $R^2$ indicates strong linear correlation, not causal direction.
2. **Comparing $R^2_{\text{adj}}$ Across Different Transformations:** You cannot compare the $R^2$ of a model predicting $Y$ directly to a model predicting $\ln(Y)$, because $\text{SST}$ is measured on fundamentally different scales.
3. **Over-reliance on $R^2_{\text{adj}}$ Alone:** Always inspect residual diagnostics (residual vs. fitted plots, Q-Q plots) to ensure linearity, normality, and homoscedasticity assumptions hold.

## 8. Summary Diagnostic Flowchart

```text
[Fit OLS Regression Model]
           │
           ▼
[Check Overall F-Test]
  ├── p >= 0.05 ───► Model does not explain variance beyond noise. Stop / Respecify.
  └── p < 0.05 ────► Model explains significant variance. Proceed:
                           │
           ┌───────────────┴──────────────┐
           ▼                              ▼
    [Assess Fit]                  [Inspect Coefficients]
    • Check R² & Adj R²           • Examine t-tests & p-values
    • Check Root MSE (s_e)        • Check signs & magnitude
                                  • Check VIF for multicollinearity
                                          │
                                          ▼
                               [Validate Assumptions]
                               • Residual vs Fitted (Linearity / Homoscedasticity)
                               • Normal Q-Q Plot (Normality of residuals)
                               • Leverage / Cook's D (Influential outliers)
```