# Cheatsheets

If you forget ggplot2 syntax, here is a cheatsheet for your reference:
[ggplot2 Cheatsheet](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf)

# Question Set 1


*Background: Differentially expressed genes results [Diff_Expression_results.tsv](https://github.com/StatBiomed/BMDS-book/raw/main/notebooks/chapter1-R/Diff_Expression_results.tsv) (or [view here](https://github.com/StatBiomed/BMDS-book/blob/main/notebooks/chapter1-R/Diff_Expression_results.tsv)) to this folder. Check your current working directory using the `getwd()` function and change it to the newly created folder.

*Hint*: Use `setwd("<path to R_exercises>")` to change the working directory, replacing `<path to R_exercises>` with the actual path to the folder, or use the `Session` menu in RStudio to set the working directory.

**What is Differential Expression?**

Every cell in your body has the same DNA, but different cells use different parts of that DNA to do their jobs. Gene expression is the process where a gene is turned "on" to make proteins that keep the cell running.

Differential expression simply means comparing how active genes are in two different conditions—for example, comparing healthy tissue versus diseased tissue, or treated cells versus untreated cells.

If a gene produces more copies in the diseased state, it is up-regulated. If it produces fewer, it is down-regulated.

```{r}
library(tidyverse)
df_DEG<-read.delim("https://github.com/StatBiomed/BMDS-book/raw/main/notebooks/chapter1-R/Diff_Expression_results.tsv",stringsAsFactors=FALSE)
```

* Q1: Use the `summary()` function to get an overview of the `df_DEG` data frame.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
summary(df_DEG)
```
:::

* Q2: Create a histogram of the `log2FoldChange` column in `df_DEG` using both the basic plotting function `hist()` and the `ggplot2` package.
  For the ggplot2 plot, add a figure title `Histogram of log2FoldChange`.

```{webr}
library(ggplot2)
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
hist(df_DEG$log2FoldChange)
ggplot(df_DEG, aes(x = log2FoldChange)) +
  geom_histogram(binwidth = 5) +
  ggtitle("Histogram of log2FoldChange")
```
:::

* Q3: Create a scatter plot with the x-axis as `log10(baseMean)` and the y-axis as `log2FoldChange` from the `df_DEG` data frame. Use both the basic `plot()` function and the `ggplot2` package.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
plot(
  log10(df_DEG$baseMean), df_DEG$log2FoldChange,
  xlab = "log10(baseMean)", ylab = "log2FoldChange", main = "Scatter Plot"
)
ggplot(df_DEG, aes(x = log10(baseMean), y = log2FoldChange)) +
  geom_point() +
  ggtitle("Scatter Plot of log2FoldChange vs log10(baseMean)")
```
:::

* Q4: Manipulate the data frame by adding two columns:
  - Add a column `log2FC_clip` for clipping `log2FoldChange` to the range `[-5, +5]`.
  - Add a column `is_DE` to indicate if `padj < 0.05`.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
df_DEG$log2FC_clip <- df_DEG$log2FoldChange
df_DEG$log2FC_clip[df_DEG$log2FC_clip > 5] <- 5
df_DEG$log2FC_clip[df_DEG$log2FC_clip < -5] <- -5
df_DEG$is_DE <- df_DEG$padj < 0.05
```
:::

* Q5: Use the `table()` function to count the values in the `is_DE` column.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
table(df_DEG$is_DE)
```
:::

* Q5: Using ggplot2, create a scatter plot colored by the newly added column `is_DE`.
  To make the figure clear, set the color legend title to `Is differentially expressed` and map the labels from `TRUE` to `Yes` and `FALSE` to `No`.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
ggplot(df_DEG, aes(x = log10(baseMean), y = log2FC_clip, color = is_DE)) +
  geom_point() +
  scale_color_discrete(name = "Is differentially expressed", labels = c("Yes", "No"))
```
:::

* Q6: For the scatter plot from Q5, reverse the color order so that `TRUE (Yes)` is blue and `FALSE (No)` is red.
  How do you adjust the code?
  *Hint*: ggplot uses the order of factor levels to render the color. Use `factor()` and set the `levels` parameter to control the order of categories.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
df_DEG$is_DE <- factor(df_DEG$is_DE, levels = c(TRUE, FALSE))
ggplot(df_DEG, aes(x = log10(baseMean), y = log2FC_clip)) +
  geom_point(aes(color = is_DE)) +
  scale_color_discrete(name = "Is differentially expressed", labels = c("Yes", "No"))
```
:::

* Q7: Save the scatter plot from Q8 or Q9 into a file named "My_DEG_results.pdf".

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
p <- ggplot(df_DEG, aes(x = log10(baseMean), y = log2FC_clip)) +
  geom_point(aes(color = is_DE)) +
  scale_color_discrete(name = "Is differentially expressed", labels = c("Yes", "No"))
ggsave("My_DEG_results.pdf", p)
```
:::

# Question Set 2

This question set focuses on clinical data manipulation using the `birthwt` dataset, including converting binary risk indicators to factors, implementing Leave-One-Out Cross-Validation (LOOCV) for disease/risk prediction, calculating logistic regression probabilities, and evaluating maternal-infant health risk factors.

First, we load the MASS package and the birthwt data:
*Hint*: If you never installed `MASS` before, you need to execute `install.packages("MASS")` at your console

```{r}
library(MASS)
data <- birthwt
head(data)
```

* Q1: To make the clinical data more interpretable, please change the column low to a factor such that 0 refers to normal weight and 1 refers to low birth weight.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
data$low <- factor(data$low, labels = c("normal", "low"))
```
:::

About Leave-One-Out Cross-Validation (LOOCV) in Clinical Prediction
LOOCV is a rigorous way to test how well a predictive model generalizes to unseen patient data.

Think of your data as a cohort of patients: In the birthwt dataset, you have records for 189 mothers and their newborns, where each row represents a clinical case containing maternal attributes (age, weight) and the birth weight outcome (low).

Leave one patient out at a time: In LOOCV, you set aside one patient's record as your "test" data.

Train on the remaining cohort: The remaining 188 patients become your "training" data to fit the logistic regression model.

Iterate across the cohort: You repeat this process 189 times—once for every patient—so that every single patient serves as the test case exactly once.

* Q2: Please use the first 188 rows as training data and the 189th row as testing data. Please implement the `train_n_predict(train_data, test_data)` function which use train_data to make a logistic regression model `low ~ age + lwt + ptl + ht` stored in the variable `model`, and return the prediction results on the test_data.

*Hint*: As `Low` and `High` is a dichotomous outcome. You need to use a suitable regression methods

```{webr}
train_n_predict <- function(train_data, test_data) {
    # Input:
    # train_data: birthwt data for training the model
    # test_data: birthwt data for testing the model
    # Output:
    # Numerical value: predicted probability of a low birth weight infant

        return(predict(model, newdata = test_data, type = "response"))
    }
```



::: {.hide}
**Answer Key**
```{webr}
train_n_predict <- function(train_data, test_data) {
    # Input:
    # train_data: birthwt data for training the model
    # test_data: birthwt data for testing the model
    # Output:
    # Numerical value: predicted probability of a low birth weight infant

    model <- glm(low ~ age + lwt + ptl + ht, data = train_data, family = "binomial")
    return(predict(model, newdata = test_data, type = "response"))
}
```
:::


Call the function train_n_predict below to calculate the predicted probability of low birth weight, and save it in data$predict[189].


```{webr}
## Your code starts here

```


::: {.hide}
**Answer Key**
```{webr}
data$predict[189] <- train_n_predict(data[-189, ], data[189, ])
data$predict[189]
```
:::


* Q3:

Now, we need to do this for a total of 189 times:

1.
For the first time, take out the first row as test data and use the rest of the 188 rows as training data.

2.
For the second time, take out the second row as test data and use the rest of the 188 rows as training data.

3.
And so on, for all 189 entries.

and remember to store the test result in `data$predict` (i.e., if the test data is the 1st row of the data, then `data$predict[1] <- train_n_predict(data[-1,], data[1,])`).

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
for (i in 1:nrow(data)) {
  data$predict[i] <- train_n_predict(data[-i, ], data[i, ])
}
```
:::

* Q4: After getting the predicted probability, we need to convert it into a clinical classification.
The function `train_n_predict` gives the probability that the infant has a low birth weight (low = 1).
Therefore, if `data$predict > 0.5`, we can classify the case as low birth weight, and if `data$predict < 0.5`, we classify it as normal weight.

Based on this idea, create a new column `data$p_class` with factor data type, where if `data$predict >= 0.5` it shall be low (original label for low birth weight) and normal if `data$predict < 0.5`.


```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
data$p_class <- factor(data$predict >= 0.5, labels = c("normal", "low"))
```

Noted that the labels order follow the factor levels order.  By default `factor(data$predict >= 0.5)` levels for this vector is `c(FALSE,TRUE)`, so the labels will match the order accordingly (i.e. `FALSE` -> `normal`,  `TRUE` -> `low` )

Alternatively, loop can tackle the problem as well as below

```{webr}
# Initialize the factor column first
data$p_class <- factor(rep(NA, nrow(data)), levels = c("normal", "low"))

for (i in 1:nrow(data)) {
  if (data$predict[i] >= 0.5) {
    data$p_class[i] <- "low"
  } else {
    data$p_class[i] <- "normal"
  }
}
```

Using `for` loops for data frame row-by-row manipulation in R is generally not recommended (and is considered an anti-pattern). Because data frames in R are underlying lists of vectors, growing them dynamically or modifying elements one-by-one inside a loop causes R to frequently reallocate memory in the background, leading to severe performance bottlenecks on larger datasets. Vectorized functions (like `ifelse()`, `dplyr::case_when()`, or logical vector indexing with factor levels) are significantly faster, more memory-efficient, and easier to read.

:::

* Q5: 
Now that we have classified our predictions (data$p_class) and have the actual clinical outcomes (data$low), we want to see which predictions were correct.

Please create a new logical column named data$correct using a logical comparison (==). It should evaluate to TRUE if the predicted classification matches the actual birth weight status, and FALSE otherwise.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
data$correct <- data$p_class == data$low
head(data[, c("low", "p_class", "correct")])
```
:::

* Q6: 
Model accuracy is defined as the proportion of correct predictions out of the total number of cases.

Using your logical comparison or the correct column you just created, calculate the overall prediction accuracy of your Leave-One-Out Cross-Validation (LOOCV) model. (Hint: In R, taking the mean() of a logical vector treats TRUE as 1 and FALSE as 0, giving you the exact proportion of correct values).

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
# Method 1: Using the logical column
accuracy <- mean(data$correct)
accuracy

# Method 2: Direct vector comparison
# mean(data$p_class == data$low)
```
:::

# Question Set 3

Let's go back to the question in the tutorial 1 this time you need implement a function to solve a quadratic equation and return "There is no real roots for this quadratic equation" whenever appropriate.

$$
ax^2+bx+c=0
$$

```{webr}
quadratic_equation <- function(a, b, c) {
## Your code starts here

}

# Test cases
quadratic_equation(1, 2, 0)
quadratic_equation(1, 2, 1)
quadratic_equation(1, 2, 4)
```

:::{.hide}
**Answer key**
```{webr}
quadratic_equation <- function(a, b, c) {
    # Calculate the discriminant
    discriminant <- b^2 - 4 * a * c
    
    # Check if there are real roots
    if (discriminant < 0) {
        return("There is no real roots for this quadratic equation")
    } else {
        # Calculate the two roots using the quadratic formula
        root1 <- (-b + sqrt(discriminant)) / (2 * a)
        root2 <- (-b - sqrt(discriminant)) / (2 * a)
        return(c(root1, root2))
    }
}

# Test cases
quadratic_equation(1, 2, 0)
quadratic_equation(1, 2, 1)
quadratic_equation(1, 2, 4)
```
:::
# Question Set 4

This question set is adapted from the [NIH Bioinformatics Training and Education Program](https://bioinformatics.ccr.cancer.gov/docs/data-visualization-with-r/Practice_Exercises/L2Ex/). The original question is slightly more complex. If you are interested in more challenging questions, you can visit the website.

Let's load the Titanic dataset.

```{webr}
titanic <- read.csv("https://web.stanford.edu/class/archive/cs/cs109/cs109.1166/stuff/titanic.csv", header = TRUE)
```

| Column                 | Description                                 |
|------------------------|---------------------------------------------|
| Survived               | 0 (No) / 1 (Yes)                            |
| Pclass                 | 1 / 2 / 3 (Passenger Class)                 |
| Name                   | Passenger Name                              |
| Sex                    | Male / Female                               |
| Age                    | Age                                         |
| Siblings.Spouses.Aboard | Number of siblings/spouses aboard          |
| Parents.Children.Aboard | Number of parents/children aboard          |
| Fare                   | Fare Price                                  |

* Q1: Inspect the data using `str()`, `summary()`, and `head()`.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
str(titanic)
summary(titanic)
head(titanic)
```
:::

* Q2: Explore the relationship between the passenger's age and fare. What type of plot should you use? Write the code using `ggplot2`.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
A scatter plot:

```{webr}
ggplot(titanic) +
  geom_point(aes(x = Age, y = Fare))
```
:::

* Q3: Color the points from Q2 by `Pclass`. Remember that `Pclass` is a proxy for socioeconomic status. While the values are treated as numeric upon loading, they are categorical and should be treated as such. You will need to coerce `Pclass` into a categorical (factor) variable using `factor()` or `as.factor()`.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
ggplot(titanic) +
  geom_point(aes(x = Age, y = Fare, color = as.factor(Pclass)))
```
:::

* Q4: Modify the plot from Q3 to update the legend labels to (1 = 1st Class, 2 = 2nd Class, 3 = 3rd Class).

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{webr}
ggplot(titanic) +
  geom_point(aes(x = Age, y = Fare, color = as.factor(Pclass))) +
  scale_color_discrete(labels = c("1st Class", "2nd Class", "3rd Class"))
```
:::

