# Tutorial & Lab: Transcriptomic Dynamics with R

### Dataset Overview

You will analyze brain tissue RNA-Seq profiles from mice infected with influenza over an 8-day timecourse:

* **Source:** `https://raw.githubusercontent.com/uclouvain-cbio/WSBIM1207/master/data/rnaseq.csv`
* **Key columns:** `gene`, `sample`, `expression`, `sex`, `infection`, `time`, `tissue`, `gene_biotype`

The case is adopted from @laurent_gatto_2021_5532658 with data from @blackmore2017influenza

```{r}
library(tidyverse)

rna <- read_csv("https://uclouvain-cbio.github.io/WSBIM1207/data/rnaseq.csv")
```

---

## Part 1: Data Wrangling with `dplyr`

### Question 1.1: Data Transformation

Raw RNA-Seq count data exhibits high dynamic range and right skewness.

1.  Read the dataset directly from the URL using `read_csv()`.
2.  Inspect the `rna` dataset to determine:
    -   How many observations (rows) are present?
    -   How many columns are present?
    -   What are the data types for each column?
3.  Compute a new column `log2_expr = log2(expression + 1)` using
    `mutate()`.
4.  Retain only protein coding genes using `filter()` on
    `gene_biotype == "protein_coding"`
5.  Save the cleaned table as `rna_clean`.

```{webr}
## Your code starts here

head(rna_clean)
```

::: {.hide}
**Answer Key**
```{r}
# 2. Inspect data dimensions and structure
str(rna)

# 3 & 4. Transform, filter, and drop columns
rna_clean <- rna |>
  mutate(log2_expr = log2(expression + 1)) |>
  filter(gene_biotype == "protein_coding")

head(rna_clean)
```
:::

---

### Question 1.2: Subsetting and Summary Statistics

We focus on the ***Fcrls*** gene. In the paper, researchers found that
***Fcrls*** is down-regulated during peak infection in response to the
neuroinflammatory cascade.

1.  Filter `rna_clean` for `gene == "Fcrls"` and
    `tissue == "Cerebellum"`.
2.  Store the unaggregated filtered subset as `fcrls_cere` for
    downstream plotting and statistical modeling.
3.  Compute group-level summary statistics `fcrls_summary`: group by
    `sex` and `time`, then calculate the mean (`mean_expr`), standard
    deviation (`sd_expr`), and count (`n = n()`).



```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{r}
# 1 & 3. Create unaggregated subset
fcrls_cere <- rna_clean |>
  filter(gene == "Fcrls", tissue == "Cerebellum")

# 2. Compute group-level summary statistics
fcrls_summary <- fcrls_cere |>
  group_by(sex, time) |>
  summarize(
    mean_expr = mean(log2_expr),
    sd_expr   = sd(log2_expr),
    n         = n(),
  )

print(fcrls_summary)
```
:::

---

## Part 2: Publication-Ready Plotting with `ggplot2`

### Question 2.1: Multi-Panel Trajectory Visualization

Visual inspection reveals whether temporal changes in *Fcrls* expression
are consistent across sexes.

1.  Build a scatter plot using `fcrls_cere` mapping `time` to the x-axis
    and `log2_expr` to the y-axis.
2.  Color points by `sex`.
3.  Adding sufficient information on x- and y-axis title
4.  Style the plot: set discrete x-ticks at `c(0, 4, 8)`

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{r}
ggplot(fcrls_cere, aes(x = time, y = log2_expr, color = sex)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_x_continuous(name  = "Days Post-Infection (Time)",breaks = c(0, 4, 8)) +
  scale_y_continuous(name  = "log expression")
  
```
:::

---

### Question 2.2: Distributional Comparison Across Time Points

To inspect the spread and distribution across time points without
assuming linearity, construct a boxplot with individual observations
overlaid.

1.  Create a boxplot mapping `factor(time)` to the x-axis and
    `log2_expr` to the y-axis.
2.  Overlay individual data points using `geom_jitter()` or
    `geom_point()`.
3.  Facet by `sex`.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{r}
ggplot(fcrls_cere, aes(x = factor(time), y = log2_expr, fill = sex)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  scale_fill_manual(values = c("Male" = "#1f77b4", "Female" = "#ff7f0e")) +
  scale_x_discrete(name  = "Days Post-Infection (Time)") +
  scale_y_continuous(name  = "log expression")
```
:::

---

## Part 3: Statistical Modeling

### Question 3.1: Difference Between Time Points

We want to test whether *Fcrls* expression differs between specific time
points: - Day 0 (T0) vs. Day 4 (T4) - Day 4 (T4) vs. Day 8 (T8)

1.  What is a suitable statistical test to compare the means of these
    two independent groups?
2.  Perform the test using the appropriate R function, extract the
    $p$-value, and state your conclusion.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**

**Test Selection:** Two-sample independent Welch's $t$-test (`t.test()`) 

```{r}
# T0 vs T4 comparison
fcrls_t0_t4 <- fcrls_cere |> filter(time <= 4)
t_test_0_4  <- t.test(
  fcrls_t0_t4$log2_expr[fcrls_t0_t4$time==0], 
  fcrls_t0_t4$log2_expr[fcrls_t0_t4$time==4])
t_test_0_4

# T4 vs T8 comparison
fcrls_t4_t8 <- fcrls_cere |> filter(time >=4)
t_test_4_8  <-t.test(
  fcrls_t4_t8$log2_expr[fcrls_t4_t8$time==4], 
  fcrls_t4_t8$log2_expr[fcrls_t4_t8$time==8])

t_test_4_8
```

**Conclusion:** \* Comparing Day 0 to Day 4 tests whether the acute
neuroinflammatory onset produces a significant down-regulation of
*Fcrls*. \* Comparing Day 4 to Day 8 assesses whether *Fcrls* expression
recovers as the infection begins resolving.

:::

---

### Question 3.2: Generalizing to *Apod* and Multiple Testing Correction

1.  Filter `rna_clean` for `gene == "Apod"` and
    `tissue == "Cerebellum"`, and perform the same comparison between
    Day 0 and Day 4.
2.  Can we repeat this pairwise testing procedure independently across
    all 15,000+ genes to find differentially expressed genes? If not,
    what adjustments are required?

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{r}
apod_t0_t4 <- rna_clean |>
  filter(gene == "Apod", tissue == "Cerebellum", time <= 4)

t_test_apod <- t.test(log2_expr ~ time, data = apod_t0_t4)
t_test_apod
```

**Discussion:** No, you cannot run thousands of unadjusted individual
tests. Testing thousands of genes simultaneously leads to the **multiple testing problem**: at an $\alpha = 0.05$ threshold, testing 20,000 genes
would produce roughly $20,000 \times 0.05 = 1,000$ false positives
purely by random chance.

**Required Modifications:** 

- Apply **False Discovery Rate (FDR)**
control using the Benjamini-Hochberg (BH) procedure via
`p.adjust(p, method = "BH")`. 
- In standard transcriptomic workflows, use specialized empirical Bayes shrinkage frameworks designed for count
data and small sample sizes (such as `DESeq2` or `limma-voom`).
:::

---

### Question 3.3: Stratified Linear Regression via Nested Data

Fit a simple linear regression model using lm() to investigate how Fcrls
expression changes over time across all samples in fcrls_cere

$$\text{log2\_expr} = \beta_0 + \beta_1 \cdot \text{time} + \epsilon$$

1.  Fit the linear model predicting log2_expr from time and store it in
    an object named `model_fcrls`.
2.  Inspect the model summary using `summary()`
3.  Extract the estimated slope ($\beta_1$) and check if it is
    statistically significant ($p < 0.05$).
4.  Extract the $R^2$ (coefficient of determination) value to assess
    goodness-of-fit.

```{webr}
## Your code starts here

```

::: {.hide}
**Answer Key**
```{r}
# 1. Fit the simple linear regression model
model_fcrls <- lm(log2_expr ~ time, data = fcrls_cere)

# 2. View full regression summary
summary(model_fcrls)

# 3. Extract coefficient table (slope, std error, t-value, p-value)
coef_summary <- summary(model_fcrls)$coefficients
coef_summary

# Specifically check the slope and its p-value
slope_estimate <- coef_summary["time", "Estimate"]
slope_pvalue   <- coef_summary["time", "Pr(>|t|)"]

cat("Slope (beta_1):", slope_estimate, "\n")
cat("p-value:", slope_pvalue, "\n")

# 4. Extract R-squared value
r_squared <- summary(model_fcrls)$r.squared
cat("R-squared:", r_squared, "\n")

```

**Interpretation:**

-   **Goodness-of-fit:** `r.squared` = 0.78 indicates the proportion of
    variance in `log2_expr` explained by post-infection time.
-   **Slope (**$\beta_1$): The row where `term == "time"` reports the
    slope estimate. A negative value indicates the average daily
    decrease in $\log_2$ expression over the course of the infection.

:::